---
title: "Lab 05 - Data Wrangling"
output: 
  github_document: default
  tufte::tufte_html:
    css: style.css
link-citations: yes
always_allow_html: yes
---

# Learning goals

- Use the `merge()` function to join two datasets.
- Deal with missings and impute data.
- Identify relevant observations using `quantile()`.
- Practice your GitHub skills.

# Lab description

For this lab we will be dealing with the meteorological dataset `met`. In this case, we will use `data.table` to answer some questions regarding the `met` dataset, while at the same time practice your Git+GitHub skills for this project.

This markdown document should be rendered using `github_document` document.

# Part 1: Setup a Git project and the GitHub repository

1. Go to wherever you are planning to store the data on your computer, and create a folder for this project

2. In that folder, save [this template](https://github.com/JSC370/JSC370-2024/blob/main/labs/lab05/lab05-wrangling-gam.Rmd) as "README.Rmd". This will be the markdown file where all the magic will happen.

3. Go to your GitHub account and create a new repository of the same name that your local folder has, e.g., "JSC370-labs".

4. Initialize the Git project, add the "README.Rmd" file, and make your first commit.

5. Add the repo you just created on GitHub.com to the list of remotes, and push your commit to origin while setting the upstream.

Most of the steps can be done using command line:

```sh
# Step 1
cd ~/Documents
mkdir JSC370-labs
cd JSC370-labs

# Step 2
wget https://raw.githubusercontent.com/JSC370/JSC370-2024/main/labs/lab05/lab05-wrangling-gam.Rmd
mv lab05-wrangling-gam.Rmd README.Rmd
# if wget is not available,
curl https://raw.githubusercontent.com/JSC370/JSC370-2024/main/labs/lab05/lab05-wrangling-gam.Rmd --output README.Rmd

# Step 3
# Happens on github

# Step 4
git init
git add README.Rmd
git commit -m "First commit"

# Step 5
git remote add origin git@github.com:[username]/JSC370-labs
git push -u origin master
```

You can also complete the steps in R (replace with your paths/username when needed)

```r
# Step 1
setwd("~/Documents")
dir.create("JSC370-labs")
setwd("JSC370-labs")

# Step 2
download.file(
  "https://raw.githubusercontent.com/JSC370/JSC370-2024/main/labs/lab05/lab05-wrangling-gam.Rmd",
  destfile = "README.Rmd"
  )

# Step 3: Happens on Github

# Step 4
system("git init && git add README.Rmd")
system('git commit -m "First commit"')

# Step 5
system("git remote add origin git@github.com:[username]/JSC370-labs")
system("git push -u origin master")
```

Once you are done setting up the project, you can now start working with the MET data.

## Setup in R

1. Load the `data.table` (and the `dtplyr` and `dplyr` packages), `mgcv`, `ggplot2`, `leaflet`, `kableExtra`.

```{r packages, warnings=FALSE, message=FALSE}
library(data.table)
library(dtplyr)
library(dplyr)
library(mgcv)
library(ggplot2)
library(leaflet)
library(kableExtra)
library(tidyr)
library(webshot)
```


```{r, eval=TRUE}
fn <- "https://raw.githubusercontent.com/JSC370/JSC370-2024/main/data/met_all_2023.gz"
if (!file.exists("met_all_2023.gz"))
  download.file(fn, destfile = "met_all_2023.gz")
met <- data.table::fread("met_all_2023.gz")
```

2. Load the met data from https://github.com/JSC370/JSC370-2024/main/data/met_all_2023.gz or  (Use https://raw.githubusercontent.com/JSC370/JSC370-2024/main/data/met_all_2023.gz to download programmatically), and also the station data. For the latter, you can use the code we used during lecture to pre-process the stations data:

```{r stations-data, eval = TRUE}
# Download the data
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE, LAT, LON)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

# Removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

# Read in the met data and fix lat, lon, temp
met$lat <- met$lat/1000
met$lon <- met$lon/1000
met$wind.sp <- met$wind.sp/10
met$temp <- met$temp/10
met$dew.point <- met$dew.point/10
met$atm.press <- met$atm.press/10
met$rh <- 100*((112-0.1*met$temp+met$dew.point)/(112+0.9*met$temp))^8
met_cleaned <- met[temp >= 0 & temp <= 50]
```

3. Merge the data as we did during the lecture. Use the `merge()` code and you can also try the tidy way with `left_join()`

```{r}
met_dt <- merge(
  # Data
  x     = met_cleaned,      
  y     = stations, 
  # List of variables to match
  by.x  = "USAFID",
  by.y  = "USAF", 
  all.x = TRUE,      
  all.y = FALSE
)
nrow(met_dt)
```

## Question 1: Identifying Representative Stations

Across all weather stations, which stations have the median values of temperature, wind speed, and atmospheric pressure? Using the `quantile()` function, identify these three stations. Do they coincide?

```{r}
# Compute the median values of temperature, wind speed, and atmospheric pressure
# globally
national_medians <- met_dt[,
                         list(
                           temp_50=quantile(temp, prob=.5, na.rm=TRUE),
                           wind.sp_50=quantile(wind.sp, prob=.5, na.rm=TRUE),
                           atm.press_50=quantile(atm.press, prob=.5, na.rm=TRUE)
                         )]
national_medians
```

Next identify the stations have these median values.

```{r}
# Compute the median values of temperature, wind speed, and atmospheric pressure
# for each station
station_median <- met_dt[,
                         list(
                           temp_50=quantile(temp, prob=.5, na.rm=TRUE),
                           wind.sp_50=quantile(wind.sp, prob=.5, na.rm=TRUE),
                           atm.press_50=quantile(atm.press, prob=.5, na.rm=TRUE)
                         ),
                         by=.(USAFID, STATE)]

# Find which stations have the median values of temperature
station_median[, temp_dist:=abs(temp_50 - national_medians$temp_50)]
median_temp_station <- station_median[temp_dist==0]
median_temp_station[, list(USAFID, temp_50)]
  
# Find which stations have the median values of wind speed
station_median[, wind.sp_dist:=abs(wind.sp_50 - national_medians$wind.sp_50)]
median_wind.sp_station <- station_median[wind.sp_dist==0]
median_wind.sp_station[, list(USAFID, wind.sp_50)]

# Find which stations have the median values of atmospheric pressure
station_median[, atm.press_dist:=abs(atm.press_50 - national_medians$atm.press_50)]
median_atm.press_station <- station_median[atm.press_dist==0]
median_atm.press_station[, list(USAFID, atm.press_50)]

# Find the coincide
coincide <- station_median[temp_dist == 0 & wind.sp_dist == 0 & atm.press_dist == 0]
coincide
```

**Answer:** The stations are listed above. They coincide in 1 station - USAFID 723119.

Knit the document, commit your changes, and save it on GitHub. Don't forget to add `README.md` to the tree, the first time you render it.

## Question 2: Identifying Representative Stations per State

Now let's find the weather stations by state with closest temperature and wind speed based on the euclidean distance from these medians. 

```{r, warning=FALSE}
# Compute the median values of temperature, wind speed, and atmospheric pressure
# for each state
state_medians <- met_dt[,
                         list(
                           state_temp_50=quantile(temp, prob=.5, na.rm=TRUE),
                           state_wind.sp_50=quantile(wind.sp, prob=.5, na.rm=TRUE)
                         ),
                         by=.(STATE)]

# Compute the euclidean distance from each station median to the median of the state
station_euc_state <- left_join(station_median, state_medians, by = "STATE")
station_euc_state <- station_euc_state[, 
                                       state_euc_dist := 
                                         sqrt((temp_50 - state_temp_50)**2 + 
                                                (wind.sp_50 - state_wind.sp_50)**2)]

# Find the stations with minimum euclidean distance in each state
station_euc_state <- station_euc_state[,
                                      .SD[state_euc_dist == min(state_euc_dist)],
                                      by=STATE]
station_euc_state[, list(USAFID, STATE, temp_50, wind.sp_50, state_temp_50, 
                           state_wind.sp_50, state_euc_dist)]
```

Knit the doc and save it on GitHub.

## Question 3: In the Geographic Center?

For each state, identify which station is closest to the geographic mid-point (median) of the state. Combining these with the stations you identified in the previous question, use `leaflet()` to visualize all ~100 points in the same figure, applying different colors for the geographic median and the temperature and wind speed median.

```{r, warning=FALSE}
# Compute the geographic mid-point (median) of the state
state_geo_medians <- met_dt[,
                         list(
                           lon_50=quantile(lon, prob=.5, na.rm=TRUE),
                           lat_50=quantile(lat, prob=.5, na.rm=TRUE)
                         ),
                         by=.(STATE)]

# Find the location of each station
station_geo <- met_dt[, list(STATE, lon, lat), by=.(USAFID)] %>%
  unique()

# Compute the euclidean distance from each station to the geographic mid-point of the state
station_geo_euc <- left_join(station_geo, state_geo_medians, by = "STATE")
station_geo_euc <- station_geo_euc[, 
                                   geo_euc_dist := 
                                     sqrt((lon - lon_50)**2 + 
                                            (lat - lat_50)**2)]

# Find the stations closest to the geographic mid-point (median) of the state.
station_geo_euc <- station_geo_euc[,
                                   .SD[geo_euc_dist == min(geo_euc_dist)],
                                   by=STATE]
station_geo_euc
```

```{r, warning=FALSE}
# Find the locations of such stations
station_euc_state_map <- left_join(station_euc_state[, list(USAFID)], 
                                   station_geo, by = "USAFID")
station_geo_euc  <- left_join(station_geo_euc[, list(USAFID)], 
                              station_geo, by = "USAFID")

# Visualize all points in the same figure with different colors
map <- leaflet() %>%
  addTiles() %>%
  # temperature and wind speed median
  addCircleMarkers(
    data = station_euc_state_map,
    ~lon, ~lat,
    color = "red",
    radius = 3,
    popup = ~paste("USAFID: ", USAFID, ", State: ", STATE)
  ) %>%
  # geographic median
  addCircleMarkers(
    data = station_geo_euc,
    ~lon, ~lat,
    color = "blue",
    radius = 3,
    popup = ~paste("USAFID: ", USAFID, ", State: ", STATE)
  ) %>%
  addLegend(
    colors = c("red", "blue"),
    labels = c("Temperature and Wind Speed Median", "Geographic Median"),
    title = "Colors for different Medians"
  )
map
```

Knit the doc and save it on GitHub.

## Question 4: Summary Table with `kableExtra`

Generate a summary table using `kable` where the rows are each state and the columns represent average temperature broken down by low, median, and high elevation stations. 

Use the following breakdown for elevation:

- Low: elev < 93
- Mid: elev >= 93 and elev < 401
- High: elev >= 401

```{r}
# Create breakdown for elevation
met_elev <- met_dt[,
                   elev_category := case_when(
                     elev < 93 ~ "Low",
                     elev >= 93 & elev < 401 ~ "Mid",
                     elev >= 401 ~ "High",
                     TRUE ~ NA)
                  ]

# Calculate average temperature for each state and elevation category
summary_table <- met_elev[, 
                          .(avg_temp = mean(temp, na.rm = TRUE)), 
                          by = .(STATE, elev_category)]

opts <- options(knitr.kable.NA = "")
# Generate the summary STATE# Generate the summary table with kable
summary_table %>%
  spread(key = elev_category, value = avg_temp) %>%
  kable(caption = "Average Temperature by Elevation Category and State",
        digits = 2,
        align = "cccc") %>%
  kable_styling()
```


Knit the document, commit your changes, and push them to GitHub. 


## Question 5: Advanced Regression

Let's practice running regression models with smooth functions on X. We need the `mgcv` package and `gam()` function to do this.

- using your data with the median values per station, first create a lazy table. Filter out values of atmospheric pressure outside of the range 1000 to 1020. Examine the association between temperature (y) and atmospheric pressure (x). Create a scatterplot of the two variables using ggplot2. Add both a linear regression line and a smooth line.

- fit both a linear model and a spline model (use `gam()` with a cubic regression spline on atmospheric pressure). Summarize and plot the results from the models and interpret which model is the best fit and why.

```{r}
# Create a lazy table
met_lz <- lazy_dt(station_median, immutable = FALSE)

# Filter out values of atmospheric pressure outside of the range 1000 to 1020
met_lz <- met_lz %>%
  filter(atm.press_50 >= 1000 & atm.press_50 <= 1020)

# Create a scatterplot of temperature (y) and atmospheric pressure (x) using ggplot2.
ggplot(as.data.table(met_lz), aes(x = atm.press_50, y = temp_50)) +
  geom_point() +
  geom_smooth(method = "lm", 
              formula = y ~ x,
              se = FALSE, 
              color = "blue") +
  geom_smooth(method = "gam", 
              formula = y ~ s(x, bs="cr"),
              se = FALSE, 
              color = "red") +
  labs(x = "Atmospheric Pressure", y = "Temperature")

# Fit a linear model
linear_model <- lm(temp_50 ~ atm.press_50, data = met_lz)
# Summarize linear model
summary(linear_model)

# Fit a spline model using gam
spline_model <- gam(temp_50 ~ s(atm.press_50, bs="cr"), data = as.data.table(met_lz))
# Summarize spline model
summary(spline_model)

# Plot results from both models
plot(spline_model)
par(mfrow=c(2,2))
plot(linear_model)

```

**Answer:**
- From the scatterplot, we can see that both the linear regression line and the smooth line are going down as atmospheric pressure increases with a moderate absolute slope. This indicates that as atmospheric pressure increases, temperature tends to go down. There is a moderate negative relationship between temperature and atmospheric pressure.

- For the fitted linear model, the p-value < 2.2e-16, which is very small, indicating that we have strong evidence against the null hypothesis. Additionally, the fitted slope is -1.14078, indicating that as atmospheric pressure increases by 1, temperature goes down by 1.14 °C on averages. There is a linear negative relationship between temperature and atmospheric pressure. Also, the linear model has a residual standard error 3.759 and R-squared 0.2642.
- For the fitted spline model, the p-value of coefficients and smooth terms are both <2e-16, which indicates that we have strong evidence against both null hypothesis. Additionally, the estimated coefficient is 20.9400 and estimated edf is 8.666, which indicates a positive parametric coefficient and a highly non-linear relationship. Also, the spline model has a GCV 13.63 and adjusted R-squared 0.296.
- Although results from both models are significant, the adjusted R-squared for the spline model is a bit higher, which indicates better model fit in terms of explaining the variability in the data. Additionally, from the plot, we can see that the spline model captures the non-linear relationship between atmospheric pressure and temperature, providing a more flexible and potentially more accurate representation of the data compared to the linear model.
## Deliverables

- .Rmd file (this file)

- link to the .md file (with all outputs) in your GitHub repository